Estimating dark matter halo concentration using the integrated mass profile

### 2016-04-17 paper/paper.tex
comentarios juan carlos

The internal DM distribution in a halo is usually parameterized
through the density profile.  
In a first approximation this profile is spherically symmetric; the
density only depends on the radial coordinate.  
One the most popular radial parameterizations is the
Navarro-Frenk-White (NFW) profile 
\citep{NFW}.   
If one is not interested in the very central region of the
dark matter halo (where galaxy formation takes place, and where the
effects of baryon physics on dark matter distribution are still 
unknown) one can also assume that this profile is universal.
\citep{Navarro2010}.  
 This profile is a double power law in radius, where the transition
break happens at the so-called scale radius, $r_s$.  
The ratio between the scale radius and the halo virial radius $R_v$ is
known as the concentration $c=R_v/r_s$.

@@

\begin{figure}
\begin{center}
  \includegraphics[width=0.47\textwidth]{error_conc_bolshoi.pdf}
\end{center}
\vspace{-0.5cm}
\caption{Uncertainties on different concentration estimates.
  The large symbols show the standard deviation of the fractional
  offset ($f_{\rm off}$ Eq. \ref{eq:f_off}) from the downsampling
  experiments.  
  The small dots correspond to the $1-\sigma$ uncertainty on the
  concentration for each halo as estimated by the MCMC algorithm in
  used for the new method.
  Uncertainties from fluctuations on the halo particle number are
  larger than the statistical uncertainties from the fitting process. 
  \label{fig:downsampling_err}}
\end{figure}

@@

We perform first a convergence test on the N-body data to find
possible biases in the concentration estimates as a function of
particle number. 
We select $8$ halos from the simulation that have at least $3\times
10^5$ particles. 
We then downsample the total number of particles in each halo by
factors of $10, 20, 45, 100, 215, 460, 1000$ to measure its
concentration with both the integrated mass and the velocity method.
each downsampling factor we repeat the operation $100$ times.

@@

Figure \ref{fig:downsampling} summarizes our results. 
The plot shows the average value of $f_{\rm off}$ as a function
of the particle number.
As expected for large enough sample particle numbers, $N_{s}>5\times
10^3$, the results of the two algorithms coincide and $\langle f_{\rm
  off}\rangle\sim 0$.  
For lower number of particles, the results of the two algorithms
deviate from the expected concentration value.  
The velocity method overestimates by a factor $\langle f_{\rm 
  off}\rangle\sim 0.25$ the concentration values for the lowest
sample particle numbers, $N_{s}\sim 100$. 
Around the same sampling scale, the new algorithm shows a more stable
behaviour with an understimate by a factor of $\langle f_{\rm 
  off}\rangle \sim 0.02$.  
This shows that in the more complex setup of a cosmological setup our
algorithm continues to be stable below the $5\%$ level for low 
particle numbers.

@@

Figure \ref{fig:downsampling} summarizes different uncertainty
results. 
Large symbols show the standard deviation on $f_{\rm off}$
from the downsampling experiment.
Small dots show the uncertainty on each individual concentration value
as estimated by the MCMC algorithm.
The line shows the $\propto N^{-1}$  locus.
This figure demonstrates that the statistical uncertainties are at
least one order of magnitude than the uncertainty associated to
downsampling experiment.

@@

We then notice that our methods follows closely the results of the
velocity method at higher masses, $M_h>10^{12}\hMsun$ or equivalently
for $>5\times10^3$ particles. 
For masses below that threshold our method saturates at a
concentration value of $\sim 7$ while the velocity method continues to
give higher concentration values.

We hypothesize that the increase in the results for the velocity
method below $5\times 10^{4}$ particles come from the systematic
bias described in the previous subsection.  
To test the general consistency of this hypothesis wwe correct the
concentration values in the velocity and integrated mass methods by
a factor of $1/(1+f_{\rm off})$, using the definition in Equation
(\ref{eq:f_off}) and the parameters obtained from the data presented
in Figure (\ref{fig:downsampling}).   
The correction brings into perfect agreement the results between the
velocity and the integrated mass method. 
This suggests that the concentration at lower masses might very well
reach a plateu in the halo mass range $10^{10}< M_{h}/\hMsun < 10^{12}$ .

@@

>>>>

The internal DM distribution in a halo is usually parameterized
through the density profile.  In a first approximation this profile is
spherically symmetric; the density only depends on the radial
coordinate.  One {\bf of} the most popular radial parameterizations is
the Navarro-Frenk-White (NFW) profile \citep{NFW}.  If one is not
interested in the very central region of the dark matter halo (where
galaxy formation takes place, and where the effects of baryon physics
on dark matter distribution are still unknown) one can also assume
that this profile is universal.  \citep{Navarro2010}.  This profile is
a double power law in radius, where the transition break happens at
the so-called scale radius, $r_s$.  The ratio between the scale radius
and the halo virial radius $R_v$ is known as the concentration
$c=R_v/r_s$.

@@

\begin{figure}
\begin{center}
  \includegraphics[width=0.47\textwidth]{error_conc_bolshoi.pdf}
\end{center}
\vspace{-0.5cm}
\caption{Uncertainties on different concentration estimates.  The
  large symbols show the standard deviation of the fractional offset
  ($f_{\rm off}$ Eq. \ref{eq:f_off}) from the downsampling
  experiments.  The small dots correspond to the $1-\sigma$
  uncertainty on the concentration for each halo as estimated by the
  MCMC algorithm {\bf delete ``in''} used for the new method.
  Uncertainties from fluctuations on the halo particle number are
  larger than the statistical uncertainties from the fitting
  process. {\bf what is the solid line showing?}
  \label{fig:downsampling_err}}
\end{figure}

@@

We perform first a convergence test on the N-body data to find
possible biases in the concentration estimates as a function of
particle number.  We select $8$ halos from the simulation that have at
least $3\times 10^5$ particles.  We then {\bf randomly} downsample the
total number of particles in each halo by factors of $10, 20, 45, 100,
215, 460, 1000$ to measure its concentration with both the integrated
mass and the velocity method. {\bf For each} downsampling factor we
repeat the operation $100$ times.

@@

Figure \ref{fig:downsampling} summarizes our results.  The plot shows
the average value of $f_{\rm off}$ as a function of the particle
number.  As expected for large enough sample particle numbers,
$N_{s}>5\times 10^3$, the results of the two algorithms coincide and
$\langle f_{\rm off}\rangle\sim 0$.  For lower number of particles,
the results of the two algorithms deviate from the expected
concentration value.  The velocity method overestimates by a factor
$\langle f_{\rm off}\rangle\sim 0.25$ the concentration values for the
lowest sample particle numbers, $N_{s}\sim 100$.  Around the same
sampling scale, the new algorithm shows a more stable behaviour with
an understimate by a factor of $\langle f_{\rm off}\rangle \sim 0.02$.
This shows that in the more complex setup of a cosmological {\bf
  simulation} our algorithm continues to be stable below the $5\%$
level for low particle numbers.

@@

Figure \ref{fig:downsampling} {\bf Aca no iría una referencia Fig 3 en
  lugar de Fig 2?} summarizes different uncertainty results.  Large
symbols show the standard deviation on $f_{\rm off}$ from the
downsampling experiment.  Small dots show the uncertainty on each
individual concentration value as estimated by the MCMC algorithm.
The line shows the $\propto N^{-1}$ locus.  This figure demonstrates
that the statistical uncertainties are at least one order of magnitude
{\bf smaller} than the uncertainty associated to downsampling
experiment.

@@

We then notice that our {\bf method} follows closely the results of
the velocity method at higher masses, $M_h>10^{12}\hMsun$ or
equivalently for $>5\times10^3$ particles.  For masses below that
threshold our method saturates at a concentration value of $\sim 7$
while the velocity method continues to give higher concentration
values.

We hypothesize that the increase in the results for the velocity
method below $5\times 10^{4}$ particles come from the systematic bias
described in the previous subsection.  To test the general consistency
of this hypothesis {\bf we} correct the concentration values in the
velocity and integrated mass methods by a factor of $1/(1+f_{\rm
  off})$, using the definition in Equation (\ref{eq:f_off}) and the
parameters obtained from the data presented in Figure
(\ref{fig:downsampling}).  The correction brings into perfect
agreement the results between the velocity and the integrated mass
method.  This suggests that the concentration at lower masses might
very well reach a plateu in the halo mass range $10^{10}< M_{h}/\hMsun
< 10^{12}$ .

@@

{\bf So far no procedure could provide a direct way to measure the
  bias in the estimated halo concentration at low particle numbers and
  a way to use it to make corrections on the estimates. In section 4.2
  it was shown a simple method that can be used on any dataset to
  estimate the errors and correct them when estimating the
  concentration at low particle numbers.}
